Direct Entropy Measurement in a Mesoscopic Quantum System

### 2018-02-02 paper/draft.tex
quick edits to figure captions

\documentclass[twocolumn,showpacs,preprintnumbers,amsmath,amssymb,pra,aps,superscriptaddress]{revtex4-1}

@@

%%% figure 1 %%%
\begin{figure}[!]
        \includegraphics[width=1.0\columnwidth]{../figures/figure_1_no-annotation.pdf}
        \caption{\label{fig:fig1}(a) Scanning electron micrograph of a device similar to the one measured, showing electrostatic gates (blue) used to define a circuit in the 2D electron gas. Grey gates are unused (grounded).  A \SI{1}{\milli\metre}-long channel is formed between the parallel gates running vertically down the image. The channel can be heated by driving current through the constriction on the left.  A 200nm diameter quantum dot is tunnel-coupled to the right side of the channel. Occupation of the dot is monitored by a capacitively-coupled charge sensor. Squares show the locations of ohmic contacts to the 2DEG. (b) Calculated charge sensor signal showing a single transition from $N \rightarrow N+1$ electrons at two temperatures ($T_{Red} > T_{Blue}$). (c) Change in the charge sensor signal with respect to temperature. The asymmetry in peak height arises from the shift in $V_0$ due to the change in entropy, $dS$, between the $N$ and $N+1$ electron states. In the case of $dS=0$, $dg$ is anti-symmetric about $V_0$.}
\end{figure}

@@

%%% figure 2 %%%
\begin{figure}[!]
        \includegraphics[width=1.0\columnwidth]{../figures/figure_2.pdf}
        \caption{\label{fig:fig2}(a) Charge sensor data for $N=0 \rightarrow 1$ at 100 and 160mK. Temperature is set by DC current through the QPC heater. (b) Derivative of the charge sensor signal with $dT \sim 60mK$. Fits to $dg$ are shown with $\delta$ as a free parameter (solid) and $\delta=0$ (dashed) (c) Transition width, $\Theta$, as a function of mixing chamber temperature. The highlighted, linear region is where all the data discussed were measured. Fitting a straight line to this data, the lever arm $\alpha$ is calculated. (d) $\Theta$ as a function of DC current through the QPC heater. The linear fit is used to convert between $I_{heat}$ and $dT$, with $R_{QPC}$ fixed at 20$k\Omega$ (e) Change in entropy as a function of AC current through the QPC heater. Top axis shows corresponding $dT$. (f) Raw data from which $dS / k_B$ in (c), and the line cut in (b) were extracted.}
\end{figure}

@@

%%% figure 3 %%%
\begin{figure}
        \includegraphics[width=1.0\columnwidth]{../figures/figure_3.pdf}
        \caption{\label{fig:fig3}(a) Transport data through the quantum dot showing the $N=0 \rightarrow 1$ transition. The first excited state, near $V_{SD}$ = \SI{1}{\milli\electronvolt}, is well-above all other energy scales in the measurement. Dashed line shows where data in (b) are taken. (b) Fixed bias transport data ($V_{SD}$ = \SI{700}{\micro\electronvolt}) showing fits to the Zeeman splitting of the ground state energy. (c) Change in entropy, extracted from $dg$ fits, as a function of parallel field. The $N=0 \rightarrow 1$ and $2 \rightarrow 3$ are overlaid here to show the behavior is the same at low parallel field. (d) and (e) show characteristic $dg$ traces from which the data in (c) were extracted.}
\end{figure}

To confirm that the measured signal derives from the electron spin, the behavior of $dS$ is investigated as a function of parallel magnetic field. Fig. \ref{fig:fig3}c shows $dS$ as a function of applied field for the 0-1 and 2-3 electron transitions. The 2-electron ground state is a spin singlet, while the 3-electron ground state is same singlet with an additional unpaired electron, making the field-dependent behavior of these transitions identical where $g \mu_{B} B$ is less than the excited state energies. Zeeman splitting, measured directly in Fig. \ref{fig:fig2}b, lifts the spin degeneracy of the unpaired electrons and the entropy of the 1- and 3- electron ground states goes to zero.

The behavior of $dS$ as a function of magnetic field can be understood using the Gibbs entropy,
%
\begin{align}
\label{eqn:Sp}
        S &= k_B \sum_{i=\pm} p_{i}(B, T) \ln{ p_{i}(B,T) }
\end{align}
%
where $p_{i}(B, T)$ is the probability of the system being in the i$^{th}$ spin state at a given field and temperature.  For states with a single unpaired electron, $p_{i}(B, T)$ can be written as a two-level system,
%
\begin{align}
\label{eqn:p}
        p_{\pm}(B, T) &= \frac{1}{1+ e^{\mp \frac{g\mu_B B}{k_B T}}}
\end{align}
%
where $\pm$ represent the spin up/down states of the unpaired electron. Fits to Eqs. \ref{eqn:Sp} and \ref{eqn:p} are shown in Fig \ref{fig:fig3}c for the 0-1 and 2-3 electron transitions. The fits allow for peak height, g-factor, and a vertical offset as free parameters with $T$ fixed at $T_{MC}$. From these fits, $dS=0.92 \ln{2}$ and $g=-0.33$ for the 0-1 transition.  Discrepancy between $g$ and the result shown in Fig. \ref{fig:fig3}b, is understood from the change in shape of the quantum dot when tuning between the $g_{dot}$ and $g_{sens}$ measurements.

%%% figure 4 %%%
\begin{figure}
        \includegraphics[width=1.0\columnwidth]{../figures/figure_4.pdf}
        \caption{\label{fig:fig4}(a) (a) Transport data through the quantum dot showing the $N=1 \rightarrow 2$ transition. Singlet-triplet splitting is roughly \SI{300}{\micro\electronvolt} and depends on applied field. Dashed line shows where data in (b) are taken. (b) TFixed bias transport data ($V_{SD}$ = \SI{1250}{\micro\electronvolt}). The triplet level is split into T$_+$, T$_0$, and $T_{-}$ (not visible). At \SI[input-protect-tokens]{\sim 9}{\tesla} T$_+$ becomes degenerate with S. g-factor is determined using $T_0$ and $T_+$ fits. (c) Change in entropy, extracted from $dg$ fits, as a function of parallel field. As the field goes from 0 to 10T, entropy of the 2-electron state goes from $0  \rightarrow \ln{2}$, while the entropy of the 1-electron state goes from $\ln{2} \rightarrow 0$. (d), (e), and (f) show characteristic $dg$ traces from which the data in (c) was extracted.}
\end{figure}

The low field (\SI{<5}{\tesla}) data in Fig. \ref{fig:fig4}c for the 1-2 electron transition is understood as the inverse of Fig \ref{fig:fig3}c. The 2-electron ground state remains in the spin singlet state with zero entropy while 1-electron ground state goes from $k_B\ln{2}$ to $0$ due to Zeeman splitting. At high fields, the 1-electron ground state remains non-degenerate, while the 2-electron ground state becomes two-fold degenerate at \SI{9}{\tesla} where the singlet and triplet ($T_+$) states cross. The singlet-triplet crossing is seen clearly in Fig.~\ref{fig:fig4}b.

The entropy for the 1-2 transition can be fit using Eq. \ref{eqn:p} as the probabilities for the 1-electron states and the following probabilities for the 2-electron singlet-triplet states,
%
\begin{align}
\label{eqn:pst}
        p_{S/T}(B, T) &= \frac{1}{1+ e^{\mp \frac{g\mu_B B - \Delta_{ST}}{k_B T}}}
\end{align}
%
where $\Delta_{ST}$ is the singlet-triplet splitting at zero field. Fitting the curve yields $g(B{=}0T)=-0.46$, $g(B{=}9T)=-0.31$, $\Delta_{ST}(B{=}9T)$ = \SI{240}{\micro\electronvolt}, $dS(B{=}0T) = 1.04\ln{2}$, and $dS(B{=}9T) = 1.01\ln{2}$. Each of these values is consistent with data in Figs.~\ref{fig:fig4}a and b as well as previous GaAs dot measurements [cite why g should change in high field]. This fitting confirms that the $dS$ signal can be related directly to a tunable degeneracy in the system.

>>>>

\documentclass[preprint,showpacs,preprintnumbers,amsmath,amssymb,pra,aps,superscriptaddress]{revtex4-1}

@@

%%% figure 1 %%%
\begin{figure}[!]
        \includegraphics[width=1.0\columnwidth]{../figures/figure_1_no-annotation.pdf}
        \caption{\label{fig:fig1}(a) Scanning electron micrograph of a device similar to the one measured, showing electrostatic gates (blue) used to define a circuit in the 2D electron gas. Grey gates are unused (grounded). Squares show the locations of ohmic contacts to the 2DEG. An electron reservoir is formed between the parallel gates running vertically down the image with a 200nm diameter quantum dot coupled to the right side. The channel can be heated by driving current through the quantum point contact on the left. Electrons tunneling between the reservoir and dot are measured with a capacitively-coupled charge sensor. (b) Calculated charge sensor signal showing a single transition from $N \rightarrow N+1$ electrons at two temperatures ($T_{Red} > T_{Blue}$). (c) Change in the charge sensor signal with respect to temperature. The asymmetry in peak height arises from the shift in $V_0$ which relates to a change in entropy between the $N$ and $N+1$ electron states. In the case of $dS=0$, $dg$ is anti-symmetric about $V_0$.}
\end{figure}

@@

%%% figure 2 %%%
\begin{figure}[!]
        \includegraphics[width=1.0\columnwidth]{../figures/figure_2.pdf}
        \caption{\label{fig:fig2}(a) Charge sensor data for $N=0 \rightarrow 1$ at 100 and 160mK. Temperature is set by DC current through the QPC heater. (b) Lockin measurement of $dg_{sens}$ with $dT \sim 60mK$. Fits to $dg$ are shown with $\delta$ as a free parameter (solid) and $\delta=0$ (dashed) (c) Transition width, $\Theta$, as a function of mixing chamber temperature. The highlighted, linear region is where the tunnel rates are determined only by temperature broadening. The lever arm $\alpha$ is calculated by fitting a straight line to this region. (d) $\Theta$ as a function of DC current through the QPC heater. Linear fit is used to convert between $I_{heat}$ and $dT$, with $R_{QPC} = $ \SI{20}{\kilo\ohm} (e) Change in entropy as a function of AC current through the QPC heater. Top axis shows corresponding $dT$. Right axis shows shift in $V_0$ per unit temperature.}
\end{figure}

@@

%%% figure 3 %%%
\begin{figure}
        \includegraphics[width=1.0\columnwidth]{../figures/figure_3.pdf}
        \caption{\label{fig:fig3}(a) Transport data through the quantum dot showing the $N=0 \rightarrow 1$ transition. The first excited state, near $V_{SD} = $ \SI{1}{\milli\electronvolt}, is well-above all other energy scales in the measurement. Dashed line at $V_{SD}$ = \SI{700}{\micro\electronvolt} shows where data in (b) are taken. (b) Fixed bias transport data showing fits to the Zeeman splitting of the ground state energy from which $g = -0.44$ is extracted. (c) Change in entropy, determined from $dg$ fits at varying parallel magnetic field. The $N=0 \rightarrow 1$ and $2 \rightarrow 3$ data are overlaid here to show the similar behavior of these states at low field. (d) and (e) show characteristic $dg$ traces from which the data in (c) were extracted. These data points are show as large markers in (c)}
\end{figure}

To confirm that the measured signal derives from the electron spin, the behavior of $dS$ was investigated as a function of parallel magnetic field. Fig.~\ref{fig:fig3}c shows $dS$ as a function of applied field for the 0-1 and 2-3 electron transitions. The 2-electron ground state is a spin singlet, while the 3-electron ground state is same singlet with an additional unpaired electron, making the field-dependent behavior of these transitions identical where $g \mu_{B} B$ is less than the excited state energies. Zeeman splitting, measured directly in Fig. \ref{fig:fig2}b, lifts the spin degeneracy of the unpaired electrons and the entropy of the 1- and 3- electron ground states goes to zero with increasing field

The behavior of $dS$ as a function of magnetic field can be understood using the Gibbs entropy,
%
\begin{align}
\label{eqn:gibbs}
        S &= k_B \sum_{i=\pm} p_{i}(B, T) \ln{ p_{i}(B,T) }
\end{align}
%
where $p_{i}(B, T)$ is the probability of the system being in the i$^{th}$ spin state at a given field and temperature.  For states with a single unpaired electron, $p_{i}(B, T)$ can be written as a two-level system,
%
\begin{align}
\label{eqn:p}
        p_{\pm}(B, T) &= \frac{1}{1+ e^{\mp \frac{g\mu_B B}{k_B T}}}
\end{align}
%
where $\pm$ represent the spin up/down states. Fits to Eqs.~\ref{eqn:gibbs} and \ref{eqn:p} are shown in Fig \ref{fig:fig3}c for each transition. The fits allow for peak height, g-factor, and a vertical offset as free parameters with $T$ fixed at $T_{MC}$. From these fits we find $dS(B=0)=0.92 \ln{2}$ and $g=-0.33$ for the 0-1 transition.  Discrepancy between $g$ and the result shown in Fig. \ref{fig:fig3}b, is understood from the change in shape of the quantum dot when tuning between the $g_{dot}$ and $g_{sens}$ measurements [cite?]. The entropy agrees well with the expected $k_{B} \ln{2}$ at zero-field and goes to zero as expected with $B_{\parallel}$

%%% figure 4 %%%
\begin{figure}
        \includegraphics[width=1.0\columnwidth]{../figures/figure_4.pdf}
        \caption{\label{fig:fig4}(a) (a) Transport data through the quantum dot showing the $N=1 \rightarrow 2$ transition. Singlet-triplet splitting is roughly \SI{300}{\micro\electronvolt} and depends on applied field. Dashed line at $V_{SD}$ = \SI{1250}{\micro\electronvolt} shows where data in (b) are taken. (b) Fixed bias transport data in parallel field. The triplet level is split into $T_+$, $T_0$, and $T_{-}$ (not visible). At \SI[input-protect-tokens]{\sim 9}{\tesla} $T_+$ becomes degenerate with $S$. $g$ is determined using $T_0$ and $T_+$ fits. (c) Change in entropy, extracted from $dg$ fits at varying parallel field. As the field goes from 0 to 10T, entropy of the 2-electron state goes from $0  \rightarrow \ln{2}$, while the entropy of the 1-electron state goes from $\ln{2} \rightarrow 0$. Fit to this behavior (dashed) is described in the text (d), (e), and (f) show characteristic $dg$ traces from which the data in (c) were extracted. These data points are show as large markers in (c)}
\end{figure}

The low field (\SI{<5}{\tesla}) data in Fig. \ref{fig:fig4}c for the 1-2 electron transition is understood as the inverse of Fig \ref{fig:fig3}c. The 2-electron ground state remains in the spin singlet state with zero entropy while 1-electron ground state entropy goes from $k_B\ln{2}$ to zero due to Zeeman splitting. At high fields, the 1-electron ground state remains non-degenerate, while the 2-electron ground state becomes two-fold degenerate at \SI{9}{\tesla} where the singlet ($S$) and triplet ($T_+$) states cross. The singlet-triplet crossing is seen clearly in Fig.~\ref{fig:fig4}b.

The entropy for the 1-2 transition can again be fit using Eq.~\ref{eqn:gibbs}, with probabilities given by Eq.~\ref{eqn:p} for the 1-electron states and the following probabilities for the 2-electron singlet-triplet states,
%
\begin{align}
\label{eqn:pst}
        p_{S/T}(B, T) &= \frac{1}{1+ e^{\mp \frac{g\mu_B B - \Delta_{ST}}{k_B T}}}
\end{align}
%
where $\Delta_{ST}$ is the singlet-triplet splitting at zero field. Fitting the data in Fig.~\ref{fig:fig4}c we find $g(B{=}0T)=-0.46$, $g(B{=}9T)=-0.31$, $\Delta_{ST}(B{=}9T)$ = \SI{240}{\micro\electronvolt}, $dS(B{=}0T) = 1.04\ln{2}$, and $dS(B{=}9T) = 1.01\ln{2}$. Each of these values is consistent with data in Figs.~\ref{fig:fig4}a and b as well as previous GaAs dot measurements [cite why g should change in high field]. This fit confirms that the $dS$ signal can be related directly to a tunable degeneracy in the system.
